Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-09 application.tex
major updating
nearly done with major update

\subsection{Newtonian cooling}\label{newton_cool}
For $l\ll l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right). 
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{energy_eq}), we identify 
\begin{align}\label{tc_newton_cool} 
  t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
\end{align}
as the thermal relaxation timescale for perturbation lengthscales
$L\ll l_\mathrm{rad}$.

@@

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $\zeta\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)\zeta^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}\zeta^{33/7}+\frac{1}{\khat^2}\right).    
\end{align}
We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model,  
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\zeta^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is neccessary for the
fundamental VSI when $\khat\gg 1$. For $\khat\to0$, this
becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
\S\ref{iso_vsi_beta_crit}).

@@

\subsection{Numerical example}\label{mmsn_example}
Consider  the MMSN at $r=50\mathrm{AU}$. Eq. \ref{mmsn_epsilon} gives
$\epsilon \simeq 
0.067$ which implies, using Eq. \ref{bcrit_mmsn}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.072. 
\end{align*} 
To estimate a thermal relaxation timescale, we set $\khat=10$ 
in Eq. \ref{beta_mmsn} to obtain 
\begin{align*}
  \beta \simeq 0.049.
\end{align*}
(In fact, Fig. \ref{mmsn_bcrit_bcool} shows that $\beta$  is not
sensitive to $\khat$ at $r=50\mathrm{AU}$ for $\khat\gg1$.)
The MMSN is unstable to the fundamental VSI, but not
strongly.

Using the MMSN disk parameters at $r=50\mathrm{AU}$ and the
scale-dependent thermal relaxation $\beta(\khat)$ given by
Eq. \ref{beta_mmsn} in our linear code, we calculate the fundamental
VSI growth rates in Fig. \ref{mmsn_growth}. Growth rates
are weakly depedent on $\khat$ until $\khat\gtrsim50$ beyond which the
mode is rapidly stabilized. A representative growth rate is 
\begin{align*}
  \nu =0.059\epsilon\Omega_k = 4.0\times10^{-3}\Omega_k 
\end{align*}
for $\khat=10$. This corresponds to a characteristic growth time $\sim
40$ orbits.

@@

Fig. \ref{mmsn_eigenW} show several eigenfunctions for the modes in
Fig. \ref{mmsn_growth}. As $\khat$ is increased, the disturbance
become more concentrated near the vertical boundaries. Thus, modes
with very large $\khat$ may not be important in practice because they
predominantly affect the disk atmosphere, which contains little mass.

\begin{figure}
   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_mmsnkx1.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_mmsnkx10.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorW_mmsnkx50.ps} 
  \caption{Fundamental VSI modes in the MMSN at
    $r=50\mathrm{AU}$. The eigenfunction $W$ for $\khat=1$ (top),
    $\khat=10$ 
    (middle) and $\khat=50$ (bottom) are shown. The real (imaginary)
    part of $W$ is plotted as the solid (dotted)
    lines. The thermal relaxation timescale $\beta(\khat)$ is given by
    Eq. \ref{beta_mmsn} and is scale-dependent. \label{mmsn_eigenW}}    
\end{figure}
% only find one unstable mode -> close to marginal 
% larger kx

\subsubsection{Vertically-depedent thermal relaxation} 
Note that for a vertically isothermal disk and opacity $\kappa_d(T)$, 
we can restore the $z$ dependence in $\beta$ (due to the $\rho$
dependence in  $\eta$, see Eq. \ref{eta_def}) by making the transformation
$\beta(\khat)\to\beta(\khat/\sqrt{2\pi}\hat{\rho})\equiv 
\tilde{\beta}(\khat,z)$ in  Eq. \ref{real_beta}, where
$\hat{\rho}=\rho/\rho(z=0)$.  We have have repeated the example
calculation above with $\beta\to\tilde{\beta}$. We find growth rates
were reduced, but at most by $\sim 15\%$ at $\khat=1$. Eigenfunctions
were similar.

>>>>

\subsection{Newtonian cooling}\label{newton_cool}
For $l\ll l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right). 
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{lin_energy}), we identify 
\begin{align}\label{tc_newton_cool} 
  t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
\end{align}
as the thermal relaxation timescale for perturbation lengthscales
$L\ll l_\mathrm{rad}$.

@@

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $\zeta\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)\zeta^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}\zeta^{33/7}+\frac{1}{\khat^2}\right).    
\end{align}
We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model,  
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\zeta^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is neccessary for the
fundamental VSI when $\khat\gg 1$. %For $\khat\to0$, this
%becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
%\S\ref{iso_vsi_beta_crit}).

@@

\subsection{Numerical example}\label{mmsn_example}
Consider  the MMSN at $r=50\mathrm{AU}$. Eq. \ref{mmsn_epsilon} gives
$\epsilon \simeq 
0.067$ which implies, using Eq. \ref{bcrit_mmsn}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.072. 
\end{align*} 
To estimate a thermal relaxation timescale, we set $\khat=30$ 
in Eq. \ref{beta_mmsn} to obtain 
\begin{align*}
  \beta \simeq 0.049.
\end{align*}
(In fact, Fig. \ref{mmsn_bcrit_bcool} shows that $\beta$  is not
sensitive to $\khat$ at $r=50\mathrm{AU}$ for $\khat\gg1$.)
The MMSN is unstable to the fundamental VSI, but not
strongly.

In Fig. \ref{mmsn_growth}, we calculate the fundamental
VSI growth rates using the MMSN disk parameters at $r=50\mathrm{AU}$
and the scale-dependent thermal relaxation $\beta(\khat)$ given by
Eq. \ref{beta_mmsn} in our linear code.  In this case we found a 
larger vertical domain with $\zmax=7\epsilon r$ was needed to capture
modes with $\khat\gtrsim60$, which have perturbations
concentrated near the vertical boundaries. However, modes with
$\khat\lesssim 60$ were found to be insensitive to boundary
conditions.

The most unstable mode occurs at $\khat\simeq30$ with growth rate 
\begin{align*}
  \nu =0.058\epsilon\Omega_k = 3.9\times10^{-3}\Omega_k.  
\end{align*}
This corresponds to a characteristic growth time $\sim
40$ orbits. However, growth rates are weakly depedent on $\khat$ for
$10\lesssim \khat 60$ so this may be taken as a representative growth
rate of the fundamental VSI in the MMSN at $r=50\mathrm{AU}$. The
eigenfunction for $\khat=30$ is shown in Fig. \ref{mmsn_eigenW}.

@@

% Fig. \ref{mmsn_eigenW} show several eigenfunctions for the modes in
% Fig. \ref{mmsn_growth}. As $\khat$ is increased, the disturbance
% become more concentrated near the vertical boundaries. Thus, modes
% with very large $\khat$ may not be important in practice because they
% predominantly affect the disk atmosphere, which contains little mass.

\begin{figure}
   \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/eigenvectorW_mmsnkx30.ps}
  % \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  % 0cm]{figures/eigenvectorW_mmsnkx10.ps} 
  % \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  % 0cm]{figures/eigenvectorW_mmsnkx50.ps} 
  \caption{Fundamental VSI mode in the MMSN at 
    $r=50\mathrm{AU}$. The eigenfunction $W$ for $\khat=30$ is shown. The real (imaginary)
    part of $W$ is plotted as the solid (dotted)
    lines. The thermal relaxation timescale $\beta(\khat)$ is given by
    Eq. \ref{beta_mmsn} and is scale-dependent. \label{mmsn_eigenW}}    
\end{figure}
% only find one unstable mode -> close to marginal 
% larger kx

\subsubsection{Vertically-depedent thermal relaxation} 
Note that for a vertically isothermal disk and opacity $\kappa_d(T)$, 
we can restore the $z$ dependence in $\beta$ (due to the $\rho$
dependence in  $\eta$, see Eq. \ref{eta_def}) by making the transformation
$\beta(\khat)\to\beta(\khat/\sqrt{2\pi}\hat{\rho})\equiv 
\tilde{\beta}(\khat,z)$ in  Eq. \ref{real_beta}, where
$\hat{\rho}=\rho/\rho(z=0)$.  We have have repeated the example
calculation above with $\beta\to\tilde{\beta}$. 
%We find growth rates
%were reduced, but at most by $\sim 15\%$ at $\khat=1$. Eigenfunctions
%were similar.
